Studying measurement invariance in factor analysis models via stochastic processes

### 2010-11-05 MIpaper/par_instable.tex
Updated based on Z feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@12 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\abstract{Common methods for the study of measurement invariance in a
  factor-analytic context include
  likelihood ratio tests, Lagrange multiplier tests, and
  Wald tests. 
  These methods require advance definition of the number of groups, group
  membership, and offending model parameters.
In this paper, we construct tests of parameter
invariance based on stochastic processes of casewise
derivatives of the likelihood function.  These tests can be viewed as
generalizations of the Lagrange multiplier test, and they are especially
useful for: (1) isolating specific parameters affected by measurement
invariance violations, and (2) identifying subgroups of individuals
that violated measurement invariance based on a continuous auxiliary
variable.  The tests are presented and illustrated in detail, along
with simulations examining the tests' abilities in controlled conditions.}

@@

The assumption that parameters are
invariant across observations is a
fundamental tenet of many statistical models.  A specific type of 
parameter invariance, measurement invariance, has implications for the
general design and use of psychometric scales.
A set of scales are defined to be measurement invariant with respect to
an auxiliary variable $V$ if:
\begin{equation}
    \label{eq:midef}
      f({\bf X} | T, V) = f({\bf X} | T),
\end{equation}
where ${\bf X}$ is a matrix of scale scores, $T$ is the latent
variable that the scales purport to measure, and $f$ is some assumed
distributional form.  This concept is 
particularly important because violations can render the scales
useless.  
That is, if a set of scales violates the condition described by
Equation~\eqref{eq:midef}, then individuals with the same
score on $T$ can systematically receive 
different scale scores.
This may lead researchers to conclude subgroup
differences on a wide variety of interesting constructs 
when, in reality, the scales are the sole cause of the 
differences.  Further, it can be inappropriate to incorporate scales
violating measurement invariance into structural equation
models, where relationships between constructs are hypothesized.  Horn
and McArdle \citeyear{HorMca92} concisely summarize the impact of
these issues, stating ``Lack of evidence of measurement 
invariance equivocates conclusions and casts doubt on theory in the
behavioral sciences'' (p.\ 141).  Borsboom \citeyear{Bor06a} also
notes that researchers often fail to assess whether measurement
invariance holds.

Focusing on applications, measurement
invariance is commonly assessed with respect to categorical variables
such as ethnicity. 
These assessments are often carried out using nested Multiple Groups
Models (e.g., \citeNP{Jor71,Bol89}) along with Likelihood Ratio tests,
though 
Lagrange Multiplier and Wald tests may also be constructed for this
purpose.
Scales may also violate measurement invariance with
respect to a continuous variable such as income.  In this situation,
there are no groups, but the condition in Equation~\eqref{eq:midef}
may still be violated.  The situation involving continuous auxiliary
variables is directly addressed by the tests proposed in this paper.
The tests proposed in this paper also have the ability to isolate
specific model parameters that violate measurement invariance.
Focusing on this point, Millsap's \citeyear{Mil05} review of 
measurement invariance methods cites ``locating the violation of the
invariance'' as one of the major outstanding problems.  The tests
resulting from the proposed project directly address this issue,
leading to an important advance in measurement invariance research.

In the following pages, we first provide a brief review of Likelihood
Ratio Tests, Lagrange Multiplier Tests, and Wald Tests.  We then
proposed tests that generalize the Lagrange Multiplier
tests, providing background and relevant theoretical results.  Next,
we study the 
proposed tests' abilities through example and simulation.  Finally,
we discuss some interesting extensions of the tests.
Throughout the manuscript, we use the term {\em{test}} to refer to a
statistical test and the term {\em{scale}} to refer to a psychometric 
test or scale.

\section{Existing Tests of Measurement Invariance}

Measurement invariance is often studied in a factor-analytic context.
For a $p \times 1$ vector of scales  
${\bm x}$ and $m$ factors, the model is
specified as:
% Define data model via equation
\begin{equation}
    \label{eq:famod}
    {\bm x}_i = {\bm \mu} + {\bm \lambda}{\bm z}_i + {\bm e}_i,\ \ \ i=1,\ldots,n,
\end{equation}
where ${\bm \mu}$ is a $p \times 1$ vector of scale means, ${\bm
  \lambda}$ is a $p \times m$ matrix of factor loadings, ${\bm z}_i$ is
an $m \times 1$ vector of factor scores for individual $i$, and ${\bm
  e}_i$ is a $p \times 1$ vector of error terms for individual $i$.
The model states that each scale is a
linear combination of the latent factor(s) and error.  
Multivariate normality of the ${\bm z}_i$ and of the ${\bm e}_i$ are
traditionally assumed (e.g., \citeNP{LawMax71}), leading to a model
that is often estimated via Maximum Likelihood methods.  We restrict
ourselves to Maximum Likelihood estimation in this manuscript, leaving
us with at least three statistical approaches to assessing measurement
invariance: the Likelihood Ratio Test, the
Lagrange Multiplier Test, and the Wald Test.  While these three tests
are asymptotically equivalent \cite{Sat89}, each is detailed
individually below.

@@

Measurement invariance is achieved
if sets of factor analysis model parameters are equal across
subgroups of individuals 
(\citeNP{Mer93}; there are many types of measurement invariance based
on different sets of model parameters being held constant, but any of these
can be examined via the proposed tests).  If measurement invariance is
not achieved, 
then the parameters assume different values for different
subgroups.  This leads to statistical tests of measurement
invariance via the comparison of two models: a restricted model where
parameters are equal across subgroups, and a ``full'' model
where parameters are allowed to assume different values for each
subgroup.  Statistically, this comparison is often 
accomplished by a likelihood ratio test (also called a
difference test).  Let the full model parameters be denoted 
${\bm \theta} = ({\bm \theta}_1^{\prime}\ {\bm
  \theta}_2^{\prime})^{\prime}$, where we can obtain the restricted model
by placing a set of equality constraints on ${\bm \theta}_1$ described
by the vector-valued function ${\bm
  h}({\bm \theta}_1) = {\bm 0}$.  A $\chi^2$ statistic testing
measurement invariance may then be calculated via:
\begin{equation}
    \label{eq:chisq}
 \chi^2 = -2(n-1)(\log L({\bm X},
\widehat{{\bm \theta}}_c) - \log L({\bm X},\widehat{{\bm \theta}})),
\end{equation}
where ${\bm{X}}$ is an $n \times p$ data matrix, $\widehat{\bm \theta}_c$
contains Maximum Likelihood parameter estimates satisfying the
constraints ${\bm h}({\bm \theta}_1) = {\bm 0}$, and $\widehat{\bm
  \theta}$ contains Maximum Likelihood parameter estimates from the
full model.  The log-likelihood function in the above equation arises
from the multivariate normal distribution and is defined as:
\begin{equation}
    \label{eq:loglik}
    \log L( \bm{X}, \bm{\theta}) = -\frac{n}{2} \left \{ p
          \log(2\pi) + \log | {\bm{\Sigma}}({\bm{\theta}}) | +
          \text{tr}[{\bm{\Sigma}}^{-1}({\bm{\theta}})({\bm{S}} +
          ({\overline{{\bm{X}}} -
            {\bm{\mu}}({\bm{\theta}})})({\overline{{\bm{X}}} -
            {\bm{\mu}}({\bm{\theta}})})^{\prime})]\right \},
\end{equation}
where $\overline{{\bm{X}}}$ is the observed mean vector, ${\bm S}$ is
the observed covariance matrix, ${\bm{\mu}}({\bm{\theta}})$ is the
model-implied mean vector, and ${\bm{\Sigma}}({\bm{\theta}})$ is the
model-implied covariance matrix.  The test resulting from
Equation~\eqref{eq:chisq} has 
degrees of freedom equal to the difference in number of free
parameters for the two models, and it rejects the hypothesis of measurement
invariance for large values of $\chi^2$.

@@

\subsection{Lagrange Multiplier Test}
The Lagrange Multiplier test (e.g., \citeNP{Sil59,ChouBent90})
focuses on a restricted model and tests whether the equality
constraints described by ${\bm h}({\bm \theta}_1) = {\bm 0}$ should be
freed.
Estimating the restricted model, we define a vector of Lagrange
multipliers $\widehat{\bm \lambda}$ that satisfy:
\begin{equation}
    \label{eq:lagrange}
    \frac{\partial \log L}{\partial {\bm \theta}} \bigg |_{\widehat{\theta}_c} +
    \frac{\partial {\bm h}}{\partial {\bm \theta}}
    \bigg |_{\widehat{\theta}_c}^{\prime} \widehat{{\bm \lambda}} = {\bm 0}.
\end{equation}
The test itself is then based on the asymptotic distribution of 
$\widehat{{\bm \lambda}}$: if $\widehat{{\bm \lambda}}$ is sufficiently close
to ${\bm 0}$, then the constraints are deemed to hold in the
population.  The resulting test statistic, which is a function of
$\widehat{\bm \lambda}$, follows a $\chi^2$ distribution for large
samples.

@@

\subsection{Relationships to Proposed Tests}
All the tests described above require advance specification of
subgroups that potentially violate measurement invariance.  This is a
major limitation in psychometrics, where researchers
ultimately want to ensure that tests are invariant with respect to all
possible subgroups of students.  For example, there are many possible
ways to define subgroups based on a continuous auxiliary variable like
income.  Assuming two subgroups, we
could define an arbitrary cutoff separating the
poorer individuals from the richer individuals.  Alternatively, we could
consider larger numbers of subgroups with similarly-arbitrary cutoffs
defining the groups.  The ultimate subgroup definitions that we
choose could potentially impact our
conclusions about whether or not a scale is measurement invariant, 
in the same way that dichotomization of continuous variables impacts
general statistical analyses \cite{MacZha02}.
Heuristically, we would like to conduct a separate Lagrange (or
Likelihood Ratio or Wald) test for each possible definition of
subgroups, then declare the scales to be measurement invariant if they
``pass'' all of these tests.  Given the large number of possible
subgroups, however, this would generally result in problems with Type
I errors.\footnote{An alternative to this procedure involves inserting the
continuous income variable directly into the factor analysis model,
which we do not pursue due 
to the potential impact on model fit and on other parameter estimates.}

  % Instead of this, can employ parameter instability tests from
  % theoretical statistics to simultaneously examine all possible
  % groupings.
Instead of testing measurement invariance separately for each possible 
subgroup definition, we can use a single test from the family of tests
proposed here.
The proposed tests can be viewed as generalizations of the
Lagrange test \cite{And93}, 
as they focus on a restricted model and 
simultaneously test for measurement invariance across all
possible subgroup definitions.  This removes the problems associated
with error rates and generally promotes a more flexible framework
for studying measurement invariance.  As previously stated, the proposed
tests also have the ability to isolate specific model parameters violating
measurement invariance.

\section{Stochastic Processes for Measurement Invariance}
The proposed tests of measurement invariance implicitly rely on an
estimated factor analysis model (one that restricts parameters to be
invariant across subgroups).
The traditional factor analysis likelihood function 
(Equation~\eqref{eq:loglik}) 
utilizes the observed covariance matrix
${\bm{S}}$ and observed mean vector $\overline{\bm{X}}$
instead of individuals' data, but the proposed tests
require the likelihood to be evaluated for each individual.  Thus, 
we utilize the equivalent casewise likelihood:
\begin{equation}
    \label{eq:caselik}
    \log \text{L}^{\ast}({\bm{X}}, {\bm{\theta}}) = \sum_{i=1}^{N} \log \mid 
\bm{\Sigma}(\bm{\theta}) \mid + \sum_{i=1}^{N} (\bm{x}_i -
\bm{\mu}(\bm{\theta}))^{\prime} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
\bm{\mu}(\bm{\theta})).
\end{equation}
This equation is often employed for factor analysis and structural
equation models with missing data (e.g., \citeNP{Wot00}), 
in which case we would take subsets of
$\bm{\Sigma}(\bm{\theta})$ and $\bm{\mu}(\bm{\theta})$ corresponding
to the observed data on each individual $i$.

Maximization of Equation~\eqref{eq:caselik} typically involves iterative
algorithms that seek values $\widehat{{\bm{\theta}}}$ such that
$\frac{\partial}{\partial \theta_j} \log \text{L}^{\ast}({\bm{X}},
{\bm{\theta}}) \big |_{\theta_j = \widehat{\theta}_j} = 0\ \ \forall\ j$.
% insert derivatives of casewise evaluated at theta-hat = 0
Such derivatives are taken across individuals, but, using the casewise
likelihood described above, we can also 
evaluate the derivatives with respect to each individual.
If a unique value of $\widehat{{\bm{\theta}}}$ maximized
each individual likelihood, then we would have:
% Equation: \psi(y_i, \theta) = d/d theta casewise_i = 0 \forall i.
\begin{equation}
    \label{eq:casederiv}
    \psi({\bm{x}}_i, \theta_j) = \frac{\partial}{\partial\theta_j}
\left[ \log \mid 
\bm{\Sigma}(\bm{\theta}) \mid + (\bm{x}_i -
\bm{\mu}(\bm{\theta}))^{\text{T}} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
\bm{\mu}(\bm{\theta}))\right] \bigg |_{\theta_j = \widehat{\theta}_j} = 0\ \ \ \forall\ i, j.
\end{equation}
In practice, no value of $\widehat{{\bm{\theta}}}$ will maximize all
individuals' likelihoods, but, if parameter invariance is satisfied, 
we can expect the $\psi(\cdot, \cdot)$ to fluctuate around zero.  This
is the idea underlying the tests described here.

\subsection{Theory}
Theoretical work in statistics (e.g., \citeNP{HjoKon02}) has shown
that functions of the 
$\psi(\cdot, \cdot)$ follow specific stochastic processes, allowing
us to use them to generally test measurement
invariance.  
That is, letting ${\bm \theta}_i$ reflect parameter values
specifically 
associated with individual $i$, measurement invariance can be 
described by the null hypothesis:
\begin{equation}
    \label{eq:h0}
    H_0: {\bm \theta}_i = {\bm \theta}_0,\ i=1,\ldots,n,
\end{equation}
where the $i$'s are typically ordered with respect to the auxiliary
variable of interest (i.e., the variable defining subgroups for tests
of measurement invariance).  For example, if parental income was the
auxiliary variable, we would order students according to parental
income and then examine whether parameters were constant across
students.  The ordering of individuals is crucial to
the test development below, but it is relaxed later in the proposal.

Despite addressing measurement
invariance, the above $H_0$ focuses no longer on subgroups of
individuals but rather on individuals themselves.  The $H_0$ is
generally applicable to situations where parameters may vary across
observations.  For example, it has most often been examined in
longitudinal models where observations are 
ordered according to time.
%In this context, {\em{change-point
%    analyses}} have received much attention from the psychometric
%community \cite{Coh08}. % CITE BOOK
Observations may also be ordered with respect to 
continuous variables other than time, a fact that allows for the
measurement invariance tests described here.  Thus, the proposed tests
stem from seminal work on
detecting turning points in longitudinal models (e.g.,
\citeNP{BroDur75,Nyb89}).

The alternative hypothesis corresponding to the above $H_0$ is that not
all the ${\bm{\theta}}_i$ are constant (i.e., measurement invariance is
violated), though specific local alternatives may also be specified to
make the tests more powerful.  We now review some of the main results
for testing $H_0$ via stochastic processes of first derivatives; more
detailed accounts include Hjort and Koning \citeyear{HjoKon02} and
Zeileis and Hornik \citeyear{ZeiHor07}.

@@

Elements of the summation in Equation~\eqref{eq:cumscore}
can be positive or 
negative, so that the cumulative scores do not necessarily increase
with $t$.  In fact, if measurement invariance (i.e., $H_0$) holds,
then the cumulative scores should 
follow a stochastic process that fluctuates around zero.  
Zeileis and Hornik \citeyear{ZeiHor07} showed specifically that, under
$H_0$:
%which generalizes earlier results from %CITE bigger journals:
% Z+H Theorem 2/Corollary 2
\begin{equation}
    \label{eq:browbridge}
{\bm{B}}(\cdot, \widehat{{\bm{\theta}}}) = {\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}})
\overset{d}{\rightarrow} {\bm W}^{0}(\cdot), % put d over rightarrow
\end{equation}
where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed information
matrix (e.g., of the factor analysis model), ${\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}})$ reflects the $p$-dimensional
cumulative sums (one for each parameter), ${\bm W}^{0}(\cdot)$ is a
$p$-dimensional Brownian bridge, and 
$\overset{d}{\rightarrow}$ reflects convergence in distribution. %?
In words, there are $p$ cumulative score processes, one for each
model parameter.  This collection of processes follows a
multidimensional Brownian bridge as derivatives accumulate in the 
summation from individual 1 to individual $n$.
Multiplication by ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}$ 
``decorrelates'' the $p$
cumulative score processes, such that each univariate process (i.e., each
process for a single model parameter) is independent of all other
processes.

The cumulative scores resulting from Equation~\eqref{eq:browbridge}
can be arranged in an $n \times p$ matrix called an {\em{empirical
    cumulative score process}}, which we will denote as
${\bm{B}}^{\text{emp}}(\widehat{\bm{\theta}})$.
Each column of this matrix follows a univariate Brownian bridge and is
based on a single factor analysis parameter. 
To carry out a test of $H_0$, we need functions that collapse across
rows (individuals) and/or columns (parameters) of the matrix.  These
functions then lead to scalar test statistics and p-values.
The specific choice of function depends on the goals of the analyst:
the row-collapsing functions are useful for isolating subsets of
parameters that violate measurement invariance, and the
column-collapsing functions are useful for isolating subgroups of
individuals who violate measurement invariance across all parameters.
Some functions result in test
statistics whose critical values can be derived analytically, while
others require critical values to be derived via the
simulation of a Brownian bridge.

For example, if we wish to test a specific parameter for
measurement invariance, ``row-collapsing''
functions include the absolute maximum or mean of the corresponding 
column in ${\bm{B}}^{\text{emp}}(\widehat{\bm{\theta}})$
\cite{ZeiHor07}. % (see ZH07 p 7)
Conversely, if we wish to generally test 
measurement invariance across sets of model parameters,
``column-collapsing'' functions 
include the maximum norm or the squared Euclidean norm of the
corresponding row in ${\bm{B}}^{\text{emp}}(\widehat{\bm{\theta}})$
\cite{HjoKon02}.

A general, two-step strategy for testing measurement
invariance involves first collapsing over columns for each individual,
yielding a single statistic for each individual. 
These $n$ statistics (ordered according to an auxiliary variable)
can then be graphed to visually define groups that violate measurement
invariance.  Functions of these $n$ statistics can also yield scalar
test statistics.

>>>>

\abstract{The issue of measurement invariance commonly arises in
  factor-analytic contexts, with methods for assessment
  including 
  likelihood ratio tests, Lagrange multiplier tests, and
  Wald tests.  These tests all require advance definition of the number of
  groups, group 
  membership, and offending model parameters.
In this paper, we construct tests of parameter
invariance based on stochastic processes of casewise
derivatives of the likelihood function.  These tests can be viewed as
generalizations of the Lagrange multiplier test, and they are especially
useful for: (1) isolating specific parameters affected by measurement
invariance violations, and (2) identifying subgroups of individuals
that violated measurement invariance based on a continuous auxiliary
variable.  The tests are presented and illustrated in detail, along
with simulations examining the tests' abilities in controlled conditions.}

@@

The assumption that parameters are
invariant across observations is a
fundamental tenet of many statistical models.  A specific type of 
parameter invariance, measurement invariance, has implications for the
general design and use of psychometric scales.
This concept is 
particularly important because violations can render the scales
useless.  
That is, if a set of scales violates measurement invariance,
then individuals with the same ``amount'' of a latent variable 
may systematically receive 
different scale scores.
This may lead researchers to conclude subgroup
differences on a wide variety of interesting constructs 
when, in reality, the scales are the sole cause of the 
differences.  Further, it can be inappropriate to incorporate scales
violating measurement invariance into structural equation
models, where relationships between constructs are hypothesized.  Horn
and McArdle \citeyear{HorMca92} concisely summarize the impact of
these issues, stating ``Lack of evidence of measurement 
invariance equivocates conclusions and casts doubt on theory in the
behavioral sciences'' (p.\ 141).  Borsboom \citeyear{Bor06a} also
notes that researchers often fail to assess whether measurement
invariance holds.

In this paper, we consider a new family of tests for assessing
measurement invariance that have important advantages over existing
tests.  We begin by developing a general framework for the tests.
This leads to a discussion of theoretical results relevant to the
proposed tests, as well as a comparison of the proposed tests with
existing tests.  Next,
we study the 
proposed tests' abilities through example and simulation.  Finally,
we discuss some interesting extensions of the tests.
Throughout the manuscript, we use the term {\em{test}} to refer to a
statistical test and the term {\em{scale}} to refer to a psychometric 
test or scale.

\section{Framework}

The tests proposed here are generally relevant to situations where the
$p$-dimensional 
data vector $\bm{x}_i, i=1,\cdots,n$ is specified to arise from a model with
density $f(\bm{x}_i | \bm{\theta}_i)$ and log-likelihood 
$\log L(\bm{\theta}_i | \bm{x}_i)$.  
For the factor-analytic situations
considered in this paper, the $\bm{x}_i$ consist of individuals' scale
scores, and 
$f()$ and $\log L()$ arise from the multivariate normal distribution.
The density is defined as:
\begin{equation}
    \label{eq:mvndensity}
    f(\bm{x}_i | \theta_i) = \frac{1}{(2\pi)^{p/2} |
      \bm{\Sigma}(\bm{\theta}_i) |^{1/2}}\exp(-\frac{1}{2}(\bm{x}_i -
    \bm{\mu}(\bm{\theta}_i))^{\prime} \bm{\Sigma}(\bm{\theta}_i)^{-1} (\bm{x}_i -
    \bm{\mu}(\bm{\theta}_i)),
\end{equation}
where ${\bm{\mu}}({\bm{\theta}}_i)$ is the
model-implied mean vector, and ${\bm{\Sigma}}({\bm{\theta}}_i)$ is the
model-implied covariance matrix.
The casewise log-likelihood is defined as:
\begin{equation}
    \label{eq:caselik}
    \log \text{L}({\bm{x}}_i, {\bm{\theta}}_i) = %\sum_{i=1}^{N} 
\log \mid 
\bm{\Sigma}(\bm{\theta}) \mid + %\sum_{i=1}^{N} 
(\bm{x}_i -
\bm{\mu}(\bm{\theta}))^{\prime} \bm{\Sigma}(\bm{\theta})^{-1} (\bm{x}_i -
\bm{\mu}(\bm{\theta})).
\end{equation}
This equation is often summed over $n$ observations, and it is also
often employed for factor analysis and structural
equation models with missing data (e.g., \citeNP{Wot00}).  We use
the likelihood here to facilitate the proposed tests.

Within the general framework outlined above, model estimation
often proceeds by assuming that $\bm{\theta}_i = \bm{\theta}\
\forall\ i$, then choosing $\hat{\bm{\theta}}$ to satisfy the
criterion:
\begin{equation}
    \label{eq:score}
  \sum_{i=1}^{n} \psi(\bm{x}_i, \hat{\bm{\theta}})] = 0,    
\end{equation}
 where $\psi(\bm{x}_i, \bm{\theta})$ 
is a score function based on the model.  
For example, in factor analysis models estimated via Maximum
Likelihood, the score function is the first derivative of $\log
\text{L}(\bm{x}_i, \bm{\theta})$ with respect to $\bm{\theta}$.
General M-estimation methods may also be handled within this
framework, though we do not discuss them here.

\section{Tests of Measurement Invariance}

Measurement invariance specifically implies that factor analysis model
parameters are invariant across individuals.  It can be framed as the
hypothesis:
\begin{equation}
    \label{eq:h0}
    H_0: {\bm \theta}_i = {\bm \theta}_0,\ i=1,\ldots,n,
\end{equation}
signifying that the parameter vector is the same for all individuals
$i$.  
This general hypothesis is very difficult to test because $n$ is large, and
there are many ways by which the $\bm{\theta}_i$ may stray from one
another.  To circumvent this problem, researchers often measure
individuals on an auxiliary variable $V$ and divide them into groups
based on that variable.  
Assuming a continuous $V$, we might divide 
individuals into two groups based on some threshold $\eta$.
Equation~\eqref{eq:h0} then becomes:
\begin{equation}
    \label{eq:h0.cut}
    H_0: \bm{\theta}_{V < \eta} = \bm{\theta}_{V > \eta},
\end{equation}
where, i.e., $\bm{\theta}_{V < \eta}$ reflects the factor analysis
parameters for all individuals with $V < \eta$.

Once we have grouped individuals based on $V$,
factor-analytic assessments
of measurement invariance are well developed.
These assessments are often carried out using nested Multiple Groups
Models (e.g., \citeNP{Jor71,Bol89}) coupled with Likelihood Ratio tests,
though 
Lagrange Multiplier and Wald tests may also be constructed for this
purpose.    While these three tests
are asymptotically equivalent \cite{Sat89}, each is detailed
individually below.

% MAY BE USEFUL SOMEWHERE
% Focusing on the latter point, Millsap's \citeyear{Mil05} review of 
% measurement invariance methods cites ``locating the violation of the
% invariance'' as one of the major outstanding problems.  The tests
% resulting from the proposed project directly address this issue,
% leading to an important advance in measurement invariance research.

@@

Measurement invariance is achieved
if sets of factor analysis model parameters are equal across
subgroups of individuals 
(\citeNP{Mer93}; there are many types of measurement invariance based
on different sets of model parameters being held constant, but any of these
can be examined via the proposed tests).  If measurement invariance is
not achieved, 
then the parameters assume different values for different
subgroups.  This leads to statistical tests of measurement
invariance via the comparison of two models: a restricted model where
parameters are equal across subgroups, and a ``full'' model
where parameters are allowed to assume different values for each
subgroup.  Statistically, this comparison is often 
accomplished by a likelihood ratio test (also called a
difference test).  Let the full model parameters be denoted 
${\bm \theta} = ({\bm \theta}_1^{\prime}\ {\bm
  \theta}_2^{\prime})^{\prime}$, where we can obtain the restricted model
by placing a set of equality constraints on ${\bm \theta}_1$ described
by the vector-valued function ${\bm
  h}({\bm \theta}_1) = {\bm 0}$.  A $\chi^2$ statistic testing
measurement invariance may then be calculated via:
\begin{equation}
    \label{eq:chisq}
 \chi^2 = -2(n-1)(\log L({\bm X},
\widehat{{\bm \theta}}_c) - \log L({\bm X},\widehat{{\bm \theta}})),
\end{equation}
where ${\bm{X}}$ is an $n \times p$ data matrix, $\widehat{\bm \theta}_c$
contains Maximum Likelihood parameter estimates satisfying the
constraints ${\bm h}({\bm \theta}_1) = {\bm 0}$, and $\widehat{\bm
  \theta}$ contains Maximum Likelihood parameter estimates from the
full model.  The log-likelihood function in the above equation arises
from the multivariate normal distribution and is defined as:
  The test resulting from
Equation~\eqref{eq:chisq} has 
degrees of freedom equal to the difference in number of free
parameters for the two models, and it rejects the hypothesis of measurement
invariance for large values of $\chi^2$.

@@

\subsection{Lagrange Multiplier Test}
The Lagrange Multiplier test (e.g., \citeNP{Sil59,ChouBent90})
focuses on a restricted model and tests whether the equality
constraints described by ${\bm h}({\bm \theta}_1) = {\bm 0}$ should be
freed.
Estimating the restricted model, we define a vector of Lagrange
multipliers $\widehat{\bm \kappa}$ that satisfy:
\begin{equation}
    \label{eq:lagrange}
    \frac{\partial \log L}{\partial {\bm \theta}} \bigg |_{\widehat{\theta}_c} +
    \frac{\partial {\bm h}}{\partial {\bm \theta}}
    \bigg |_{\widehat{\theta}_c}^{\prime} \widehat{{\bm \kappa}} = {\bm 0}.
\end{equation}
The test itself is then based on the asymptotic distribution of 
$\widehat{{\bm \kappa}}$: if $\widehat{{\bm \kappa}}$ is sufficiently close
to ${\bm 0}$, then the constraints are deemed to hold in the
population.  The resulting test statistic, which is a function of
$\widehat{\bm \kappa}$, follows a $\chi^2$ distribution for large
samples.

@@

\subsection{Extensions to Proposed Tests}
For continuous $V$, the three tests described above require the
threshold $\eta$ to be known.
  This is a
major limitation in psychometrics, where $\eta$ is often unknown or
undefined.
For example, if $V$ represents yearly income, there are many possible
values of 
$\eta$ that could be used to divide individuals into poorer and richer
groups.  The ultimate $\eta$ that we
choose could potentially impact our
conclusions about whether or not a scale is measurement invariant, 
in the same way that dichotomization of continuous variables impacts
general statistical analyses \cite{MacZha02}.

Instead of choosing a specific $\eta$, we may elect to test
individuals across all possible values of $\eta$.  For example, we
could conduct a Likelihood Ratio Test for every possible
dichotomization of $V$, yielding a test statistic for each
dichotomization.  The distribution of the maximum of these test
statistics is derived in Andrews \citeyear{And93}, providing a test of
measurement invariance over all values of $\eta$.  
This is a special case of the family of tests proposed here.  The
tests are designed for situations where
$\eta$ is unknown or where a dichotomization of $V$ is insufficient to
detect measurement invariance violations, and they can be applied both
to individual model parameters and sets of model parameters.

We now describe some general ideas for capturing measurement
invariance violations with respect to a continuous auxiliary variable
$V$.

\section{Stochastic Processes for Measurement Invariance}
The proposed tests of measurement invariance implicitly rely on an
estimated factor analysis model (one that restricts parameters to be
invariant across subgroups).  For the purposes of this paper, we
assume that the model is estimated via Maximum Likelihood.  
The factor analysis log-likelihood may be obtained by summing 
Equation~\eqref{eq:caselik} across individuals.  
Maximization of the likelihood then involves iterative 
algorithms that seek values $\widehat{{\bm{\theta}}}$ to satisfy
Equation~\eqref{eq:score}, where 
$\psi(\bm{x}_i, \theta_j) = \frac{\partial}{\partial \bm{\theta}} \log
\text{L}({\bm{x}}_i, 
{\bm{\theta}}) \big |_{\bm{\theta} = \widehat{\bm{\theta}}}$.
% insert derivatives of casewise evaluated at theta-hat = 0
Such derivatives are taken across individuals, but, using individual
terms of the log-likelihood (i.e., Equation~\eqref{eq:caselik}),
we can also 
evaluate the derivatives with respect to each individual.
If a unique value of $\widehat{{\bm{\theta}}}$ maximized
each individual likelihood, then we would have:
% Equation: \psi(y_i, \theta) = d/d theta casewise_i = 0 \forall i.
\begin{equation}
    \label{eq:casederiv}
    \psi({\bm{x}}_i, \bm{\theta}) = \frac{\partial}{\partial \bm{\theta}} \log
\text{L}({\bm{x}}_i, 
{\bm{\theta}}) \bigg |_{\bm{\theta} = \widehat{\bm{\theta}}} = \bm{0}\ \ \ \forall\ i.
\end{equation}
In practice, no value of $\widehat{{\bm{\theta}}}$ will maximize all
individuals' likelihoods, but, if measurement invariance is satisfied, 
we can expect the $\psi(\cdot, \cdot)$ to fluctuate around zero.  This
is the idea underlying the tests described here.

\subsection{Theory}
Theoretical work in statistics (e.g.,
\citeNP{BroDur75,HjoKon02,Nyb89}) has shown 
that functions of the 
$\psi(\cdot, \cdot)$ follow specific stochastic processes, allowing
us to use them to generally test measurement
invariance.  
We rely on the null hypothesis specified in Equation~\eqref{eq:h0},
which focuses on individuals instead of groups of individuals.
We now review some of the main results
for testing this $H_0$; more
detailed accounts include Hjort and Koning \citeyear{HjoKon02} and
Zeileis and Hornik \citeyear{ZeiHor07}.

@@

Elements of the summation in Equation~\eqref{eq:cumscore}
can be positive or 
negative, so that the cumulative scores do not necessarily increase
with $t$.  In fact, if measurement invariance (i.e., $H_0$) holds,
then the cumulative scores should 
follow a stochastic process that fluctuates around zero.  
Zeileis and Hornik \citeyear{ZeiHor07} showed specifically that, under
$H_0$:
\begin{equation}
    \label{eq:browbridge}
{\bm{B}}(\cdot, \widehat{{\bm{\theta}}}) = {\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}{\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}})
\overset{d}{\rightarrow} {\bm W}^{0}(\cdot), % put d over rightarrow
\end{equation}
where ${\bm{I}}(\widehat{{\bm{\theta}}})$ is the observed information
matrix (e.g., of the factor analysis model), ${\bm{W}}_n(\cdot,\widehat{{\bm{\theta}}})$ reflects the $p$-dimensional
cumulative sums (one for each parameter), ${\bm W}^{0}(\cdot)$ is a
$p$-dimensional Brownian bridge, and 
$\overset{d}{\rightarrow}$ reflects convergence in distribution.
In words, there are $p$ cumulative score processes, one for each
model parameter.  This collection of processes follows a
multidimensional Brownian bridge as derivatives accumulate in the 
summation from individual 1 to individual $n$.
Multiplication by ${\bm{I}}(\widehat{{\bm{\theta}}})^{-1/2}$ 
``decorrelates'' the $p$
cumulative score processes, such that each univariate process (i.e., each
process for a single model parameter) is unrelated to all other
processes (and is asymptotically independent of the other processes).

The cumulative scores resulting from Equation~\eqref{eq:browbridge}
can be arranged in an $n \times p$ matrix called an {\em{empirical
    cumulative score process}}, which we will denote as
${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$.
Each column of this matrix follows a univariate Brownian bridge and is
based on a single factor analysis parameter. 
To carry out a test of $H_0$, we need functions that collapse across
rows (individuals) and/or columns (parameters) of the matrix.  
These
functions then lead to scalar test statistics and p-values.
The specific choice of function depends on the goals of the analyst:
the row-collapsing functions are useful for isolating subsets of
parameters that violate measurement invariance, and the
column-collapsing functions are useful for isolating subgroups of
individuals who violate measurement invariance across all parameters.
Some functions result in test
statistics whose critical values can be derived analytically, while
others require critical values to be derived via the
simulation of a Brownian bridge.

For example, if we wish to test a specific parameter for
measurement invariance, ``row-collapsing''
functions include the absolute maximum or mean of the corresponding 
column in ${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$
\cite{ZeiHor07}. % (see ZH07 p 7)
Conversely, if we wish to generally test 
measurement invariance across sets of model parameters,
``column-collapsing'' functions 
include the maximum norm or the squared Euclidean norm of the
corresponding row in ${\bm{B}}^{\ast}(\widehat{\bm{\theta}})$
\cite{HjoKon02}.

% A general, two-step strategy for testing measurement
% invariance involves first collapsing over columns for each individual,
% yielding a single statistic for each individual. 
% These $n$ statistics (ordered according to an auxiliary variable)
% can then be graphed to visually define groups that violate measurement
% invariance.  Functions of these $n$ statistics can also yield scalar
% test statistics.
